---
title: "Assignment 2"
# format: html
format: pdf
---

# Red Wine Quality Dataset

## 1.a Reading the Dataset

```{r}
#| message: false
library(tidyverse)
library('tinytex')

red_wine_data <- read.csv("data/wine+quality/winequality-red.csv", sep=";")

# Display the first few rows to verify that the column headers
# were correctly recognized and not treated as a row of data.
# head(red_wine_data)

# Tell Quarto to render output as properly formatted table.
# First six columns
knitr::kable(
  head(red_wine_data[, 1:6]),
  caption = "First six observations of the red wine quality dataset."
)

# Last six columns
knitr::kable(
  head(red_wine_data[, 7:12]),
  caption = "First six observations of the red wine quality dataset."
)
```

## 1.b Summary Statistics

### 1. Median Quality Score

```{r}
#| include: false
median_quality <- median(red_wine_data$quality)

median_quality
```

The median quality score across all wine samples is **`r median_quality`**.

### 2. Mean Alcohol Level

```{r}
#| include: false
mean_alcohol <- mean(red_wine_data$alcohol)

mean_alcohol
```

The mean alcohol level across all wine samples is **`r mean_alcohol`**.

## 1.c Density and Volatile Acidity

```{r}
# Calculate the linear regression model
model <- lm(volatile.acidity ~ density, data = red_wine_data)

# Extract slope and R-squared
slope <- coef(model)["density"]
r_squared <- summary(model)$r.squared

ggplot(red_wine_data, aes(x = density, y = volatile.acidity)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship Between Density and Volatile Acidity",
    subtitle = paste(
      "Slope =", round(slope, 3),
      "| R² =", round(r_squared, 3)
    ),
    x = "Density",
    y = "Volatile Acidity"
  )
```

The best-fit line has a slope of **`r round(slope, 3)`**, indicating a slight positive relationship between density and volatile acidity. However, With an R² value of **`r round(r_squared, 3)`**, there is essentially no linear relationship between density and volatile acidity in this dataset.