---
title: "Exercise 1. Red Wine Quality Dataset"
format: html
---

# Red Wine Quality Dataset

## 1.a Reading the Dataset

Read the dataset into an R object. Make sure the column headers are correct, and show the first couple rows.

```{r}
#| message: false

library(tidyverse)

red_wine_data <- read.csv(
  "data/wine+quality/winequality-red.csv",
  sep = ";"
)

head(red_wine_data)
```

## 1.b Summary Statistics

```{r}
median_quality <- median(red_wine_data$quality)
mean_alcohol <- mean(red_wine_data$alcohol)

cat("Median Quality Score:", median_quality, "\n")
cat("Mean Alcohol Level:", mean_alcohol, "\n")
```

### 1. Median Quality Score: **`r median_quality`**

### 2. Mean Alcohol Level: **`r mean_alcohol`**

## 1.c Density and Volatile Acidity

Create a scatterplot showing the relationship between density and volatile acidity.

```{r}
#| message: false

# Calculate the linear regression model
model <- lm(volatile.acidity ~ density, data = red_wine_data)

# Extract slope and R-squared
slope <- coef(model)["density"]
r_squared <- summary(model)$r.squared

ggplot(red_wine_data, aes(x = density, y = volatile.acidity)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship Between Density and Volatile Acidity",
    subtitle = paste(
      "Slope =", round(slope, 3),
      "| R² =", round(r_squared, 3)
    ),
    x = "Density",
    y = "Volatile Acidity"
  )
```

The best-fit line has a slope of **`r round(slope, 3)`**, which indicates a positive relationship between density and volatile acidity. However, with an R² value of **`r round(r_squared, 3)`**, the relationship is very weak. There is essentially no relationship between the two variables.

## 1.d Alcohol Level Categories

Classify each wine sample according to the following conditions:

- **High**: Alcohol level is greater than 10.5
- **Medium**: Alcohol level is less than or equal to 10.5.

```{r}
# Create the ALevel categorical variable
red_wine_data$ALevel <- ifelse(
  red_wine_data$alcohol > 10.5,
  "High",
  "Medium"
)

# Calculate the sulphates-to-chlorides ratio
red_wine_data$sulphates_to_chlorides <-
  red_wine_data$sulphates / red_wine_data$chlorides

# Create side-by-side boxplots
ggplot(
  red_wine_data,
  aes(x = ALevel, y = sulphates_to_chlorides)
) +
  geom_boxplot() +
  labs(
    title = "Sulphates-to-Chlorides Ratio by Alcohol Level",
    x = "Alcohol Level",
    y = "Sulphates-to-Chlorides Ratio"
  )

# Count the number of High ALevel samples
high_count <- sum(red_wine_data$ALevel == "High")
```

The **High** group has a higher median sulphates-to-chlorides ratio than the **Medium** group. The **High** group also shows greater variability. Both groups contain several high-level outliers. There are **`r high_count`** wine samples in the **High** ALevel category.

## 1.e Citric Acid Distribution

Create a histogram showing the citric acid values for both the **High** and **Medium** groups. If both groups are shown in the same plot, include a legend to distinguish them.

```{r}
ggplot(
  red_wine_data,
  aes(x = citric.acid, fill = ALevel)
) +
  geom_histogram(
    bins = 20,
    alpha = 0.5,
    position = "identity"
  ) +
  labs(
    title = "Citric Acid Distribution by Alcohol Level",
    x = "Citric Acid",
    y = "Number of Wine Samples",
    fill = "Alcohol Level"
  )
```

## 1.f Additional Exploration

Continue exploring the dataset by creating two additional plots. For each plot:

- State the question or hypothesis.
- Include the code and resulting plot.
- Summarize what you observe.

### Plot 1: Alcohol Level and Wine Quality

**Question:** Does alcohol level vary across different wine qualities?

```{r}
ggplot(red_wine_data, aes(x = factor(quality), y = alcohol)) +
  geom_boxplot() +
  labs(
    title = "Alcohol Level by Wine Quality",
    x = "Wine Quality Score",
    y = "Alcohol Level"
  )
```

Higher-quality wines generally have higher alcohol levels. The median alcohol level increases noticeably for quality scores 6–8 compared with scores 3–5.

### Plot 2: pH and Fixed Acidity

**Question:** Is there a relationship between fixed acidity and pH?

```{r}
#| message: false

model <- lm(pH ~ fixed.acidity, data = red_wine_data)
slope <- coef(model)["fixed.acidity"]
r_squared <- summary(model)$r.squared

ggplot(red_wine_data, aes(x = fixed.acidity, y = pH)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Relationship Between Fixed Acidity and pH",
    subtitle = paste(
      "Slope =", round(slope, 3),
      "| R² =", round(r_squared, 3)
    ),
    x = "Fixed Acidity",
    y = "pH"
  )
```

The scatterplot shows a moderate negative linear relationship between fixed acidity and pH. The slope of −0.061 indicates that pH decreases as fixed acidity increases, while the R² value of 0.466 indicates that fixed acidity explains approximately 46.6% of the variation in pH.