---
title: "exercise2_html"
format: html
---

# Bike Sharing Dataset

## 2.a Variable Types 
Identify which predictors are quantitative and which are qualitative.

```{r}
#| output: false

library(tidyverse)

day_data <- read.csv(
  "data/bike+sharing+dataset/day.csv",
  sep = ","
)

hour_data <- read.csv(
  "data/bike+sharing+dataset/hour.csv",
  sep = ","
)

# Variables that are numerically encoded but represent categories
day_categorical <- c(
  "season", "yr", "mnth", "holiday",
  "weekday", "workingday", "weathersit"
)

hour_categorical <- c(
  "season", "yr", "mnth", "hr", "holiday",
  "weekday", "workingday", "weathersit"
)

# Quantitative variables: numeric variables that are not categorical
day_quantitative <- names(day_data)[
  sapply(day_data, is.numeric) &
  !(names(day_data) %in% day_categorical)
]

hour_quantitative <- names(hour_data)[
  sapply(hour_data, is.numeric) &
  !(names(hour_data) %in% hour_categorical)
]

# Qualitative variables: character/factor variables + numerically encoded categories
day_qualitative <- names(day_data)[
  sapply(day_data, function(x) is.character(x) | is.factor(x)) |
  names(day_data) %in% day_categorical
]

hour_qualitative <- names(hour_data)[
  sapply(hour_data, function(x) is.character(x) | is.factor(x)) |
  names(hour_data) %in% hour_categorical
]

head(day_data)
head(hour_data)

# Print results
cat("DAY DATASET\n")
cat("Quantitative variables:\n")
cat(paste(day_quantitative, collapse = ", "))
cat("\n\nQualitative variables:\n")
cat(paste(day_qualitative, collapse = ", "))

cat("\n\n\nHOUR DATASET\n")
cat("Quantitative variables:\n")
cat(paste(hour_quantitative, collapse = ", "))
cat("\n\nQualitative variables:\n")
cat(paste(hour_qualitative, collapse = ", "))

```
We can automatically check the R storage type from the variable's statistical type. If the data type is not numeric, then the variable is not quantitative. However, a qualitative (categorical) variable can still have a numeric data type. The season, year, month, holiday, weekday, workingday, and weathersit variables are all categorical, but stored as the integer data type. These variable types were hard-coded.

## 2.b Summary Statistics
For each quantitative predictor, calculate and report:

- Range
- Mean
- Standard deviation

Then determine which season has the highest average bike rental court.

```{r}
#| output: true

# Summary statistics for quantitative predictors
day_summary <- day_data %>%
  summarise(
    across(
      all_of(day_quantitative),
      list(
        Range = ~ max(.x) - min(.x),
        Mean = ~ mean(.x),
        SD = ~ sd(.x)
      )
    )
  )

day_summary

# Average bike rental count by season
season_summary <- day_data %>%
  group_by(season) %>%
  summarise(
    average_rentals = mean(cnt)
  ) %>%
  arrange(desc(average_rentals))

season_summary

highest_season <- season_summary %>%
  slice_max(average_rentals, n = 1)

cat(
  "Season with the highest average bike rental count:",
  highest_season$season,
  "\nAverage rental count:",
  round(highest_season$average_rentals, 2)
)
```
## 2.c Rentals by Weather Condition
Create boxplots showing bike rental count for each weather condition.

```{r}
# Boxplots of bike rental count by weather condition
ggplot(day_data, aes(x = factor(weathersit), y = cnt)) +
  geom_boxplot() +
  labs(
    x = "Weather Condition",
    y = "Bike Rental Count"
  ) +
  theme_minimal()

weather_summary <- day_data %>%
  group_by(weathersit) %>%
  summarise(
    median_rentals = median(cnt)
  ) %>%
  arrange(desc(median_rentals))

weather_summary

cat(
  "\nWeather condition with the highest median bike rental count:",
  weather_summary$weathersit[1]
)
```
The median bike rental count is highest under weather condition 1, with a median of 4,844 rentals. Weather condition 2 has a lower median of 4,040, while condition 3 has the lowest at 1,817. Overall, bike rentals tend to be higher under better weather conditions.

## 2.d Monthly Bike Rentals 
Create a bar plot showing the total number of bike rentals for each month of the year.

```{r}
# Calculate total bike rentals for each month
monthly_rentals <- day_data %>%
  mutate(month = format(as.Date(dteday), "%B")) %>%
  group_by(month) %>%
  summarise(total_rentals = sum(cnt)) %>%
  mutate(month = factor(month, levels = month.name))

monthly_rentals

ggplot(monthly_rentals, aes(x = month, y = total_rentals)) +
  geom_col() +
  labs(
    x = "Month",
    y = "Total Bike Rentals"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

highest_month <- monthly_rentals %>%
  slice_max(total_rentals, n = 1)

cat(
  "Month with the highest total number of rentals:",
  as.character(highest_month$month),
  "\nTotal rentals:",
  highest_month$total_rentals)
```

I extracted the month from the **dteday** variable, grouped the data by month, and summed the **cnt** variable within each month to calculate the total number of bike rentals. The month with the highest number of rentals is **`r `highest_month$total_rentals`**. 

## 2.e Relationships Among Variables 
Using the dataset, investigate the relationships among the predictors. Focus the correlation matrix on **quantitative predictors**: temp, atemp, hum, and windspeed. The rental-count variables (casual, registered, cnt) are response variables rather than predictors. Avoid including variables such as season, weekday, weathersit because their categories are represented by numerical codes.

```{r}
# Quantitative predictors
quantitative_predictors <- c(
  "temp",
  "atemp",
  "hum",
  "windspeed"
)

# Correlation matrix
correlation_matrix <- day_data %>%
  select(all_of(quantitative_predictors)) %>%
  cor()

cat("Correlation Matrix:\n")
correlation_matrix

# Scatterplot matrix
day_data %>%
  select(all_of(quantitative_predictors)) %>%
  pairs()
```
The strongest relationship is between **temp** and **atemp** with a correlation of 0.992, indicating a strong positive linear relationship. This is also seen in the scatterplot. **temp** and **atemp** are highly redundant.

## 2.f Predicting Bike Rental Count
We want to predict the total bike rental count using the other variables. We can identify useful predictors based on the relationships found in **2.e**. Since **tempt** and **atempt** are highly correlated with each other, using both could introduce redundancy. **temp**, **hum**, and **windspeed** may be useful predictors based on the correlation analysis. **season**, **yr**, and **weathersit** were not studied in the correlation analysis, but could also potentially be useful.
Correlation alone has limitations because it only measures linear relationships between quantitative variables. It does not capture nonlinear relationships of relationships involving categorical variables. It also does not establish causation.
